---
title: "cocktailr"
output:
  github_document:
    toc: true
    toc_depth: 3
    df_print: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
# Global knitr options for fast, reproducible README builds
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 4,
  dpi = 150,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
set.seed(1)
```

# cocktailr

Fast, reproducible *Cocktail* clustering for vegetation tables.

---

## Overview

**cocktailr** provides fast, reproducible *Cocktail* agglomerative clustering for **plots × species** presence/absence tables, using sparse matrices, exact φ (phi) coefficients from a single cross-product per round, and deterministic tie-breaking. It also computes **fuzzy species-by-node φ memberships** for every internal cluster node.

> **TL;DR:** A fast, deterministic R package to turn vegetation **plots × species** data into reproducible *Cocktail* cluster trees with exact φ coefficients.

---

## Installation

```r
# Install from GitHub
remotes::install_github("dvynokur/cocktailr")
```

---

## Quick Start

A minimal example that demonstrates a merge with **positive φ** and shows non-empty cluster results.

```{r quick-start}
library(cocktailr)

# Tiny matrix with positive association between sp1 & sp2
vm_pos <- matrix(c(
  1,1,
  1,0,
  0,0
), nrow = 3, byrow = TRUE,
dimnames = list(paste0("plot", 1:3), c("sp1","sp2")))

res_pos  <- cocktail_cluster(vm_pos, progress = FALSE)
res_pos$Cluster.height  # should be > 0 (e.g. +0.5)

# Parent clusters at φ ≥ 0.3
labs_pos <- clusters_at_cut(res_pos, phi = 0.3)
labs_pos
species_in_clusters(res_pos, labels = labs_pos)
```

---

## Typical workflow

An end-to-end example on a toy **plots × species** matrix.  
It shows classical clustering, dendrogram plotting, cluster inspection, fuzzy φ computation, and plot assignment.

```{r typical-workflow}
library(cocktailr)

# Toy plots × species matrix
vm <- matrix(c(1,0,1,0,1,0,
               0,1,0,1,0,1,
               1,1,0,0,1,0,
               0,0,1,1,0,0),
             nrow = 4, byrow = TRUE,
             dimnames = list(paste0("plot", 1:4), paste0("sp", 1:6)))

# 1) Classical Cocktail clustering
res <- cocktail_cluster(vm, progress = FALSE)

# 2) Plot dendrogram to PDF (disabled in README to avoid file output)
# plot_cocktail(res, "res.pdf", bands_phi = 0.30, palette = "rainbow", cex_labels = 1)

# 3) Parent clusters at φ ≥ 0.30
phi_cut <- 0.30
labs <- clusters_at_cut(res, phi = phi_cut)
labs

# 4) Diagnostic species from Cocktail object (binary membership)
species_in_clusters(res, labels = labs)

# 5) Fuzzy Cocktail: φ for species × all internal nodes
res_fuzzy <- cocktail_fuzzy(res, vm)

# 6) Diagnostic species from fuzzy φ matrix
species_in_clusters(res_fuzzy, labels = labs, min_phi = 0.20, top_k = 10)

# 7) Assign plots to parent groups (strict)
assign_strict <- assign_releves(
  x               = res,
  vegmatrix       = vm,
  mode            = "strict",
  phi             = phi_cut,
  cover_transform = "sqrt",
  compare         = "max",
  min_cover       = 0,
  min_group_size  = 1
)
table(assign_strict)

# 8) Assign plots using fuzzy φ weights
assign_fuzzy <- assign_releves(
  x               = res,
  vegmatrix       = vm,
  mode            = "fuzzy",
  phi             = phi_cut,
  cover_transform = "sqrt",
  compare         = "max",
  min_cover       = 0,
  phi_mode        = "thresh",
  min_group_size  = 1
)
table(assign_fuzzy)

# 9) Species for assigned (non-"not assigned") groups
labs_strict <- setdiff(names(table(assign_strict)), "not assigned")
if (length(labs_strict)) species_in_clusters(res, labs_strict)

labs_fuzzy <- setdiff(names(table(assign_fuzzy)), "not assigned")
if (length(labs_fuzzy)) species_in_clusters(res_fuzzy, labs_fuzzy, min_phi = 0.15, top_k = 20)
```

---

### (Optional) Attach assignments to header data frame

If you have a metadata table `hea` with a column `releve_number`, you can join assignments back to it.  
(Example not evaluated in README builds.)

```{r attach-assignments, eval=FALSE}
library(dplyr)
library(tibble)

assign_df <- tibble(
  plot_id         = names(assign_strict),
  cocktail_strict = unname(assign_strict),
  cocktail_fuzzy  = unname(assign_fuzzy)
)

hea2 <- hea %>%
  mutate(plot_id = as.character(releve_number)) %>%
  left_join(assign_df, by = "plot_id") %>%
  select(-plot_id)
```
---

## Reference

See function help for details:

- `?cocktail_cluster` – deterministic Cocktail clustering  
- `?plot_cocktail` – draw dendrograms  
- `?clusters_at_cut` – find parent clusters at φ cuts  
- `?cocktail_fuzzy` – compute fuzzy species × node φ  
- `?assign_releves` – assign plots (strict/fuzzy)  
- `?species_in_clusters` – list diagnostic species for clusters

---

© 2025 Denys Vynokurov & Helge Bruelheide. Licensed under MIT.
