---
title: "cocktailr"
output:
  github_document:
    toc: true
    toc_depth: 3
    df_print: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.path   = "man/figures/README-",
  fig.width  = 7,
  fig.height = 5,
  dpi        = 150,
  out.width  = "100%",
  warning    = FALSE,
  message    = FALSE,
  cache      = FALSE
)
set.seed(1)
```

# cocktailr

*Cocktail* clustering for diagnostic species-based vegetation classification.

***

## Citation

If you use `cocktailr`, please cite the GitHub repository and the original Cocktail method papers listed below. A manuscript describing the package and workflow has been prepared for submission.

## Overview

**cocktailr** provides fast and reproducible *Cocktail* clustering for vegetation-plot data. It identifies groups of co-occurring species from **plots × species** tables or **long-format vegetation tables** (plot–species–value) and supports their use in diagnostic species-based vegetation classification. 
A possible mnemonic for **cocktailr** is **CO-occurrence Clustering with Known Thresholds for Assignment, Interpretation, and Labelling in R**.

The package implements:

- **Hierarchical Cocktail clustering** of species (`cocktail_cluster()`).
- **Dendrogram plotting** with φ heights and optional cluster bands (`cocktail_plot()`).
- Extraction of **parent clusters at a φ cut** (`clusters_at_cut()`).
- **Selection of strong clusters** by a combined score (`select_clusters()`).
- **Diagnostic species lists** for clusters or combinations of clusters (`species_in_clusters()`).
- Listing **plots (relevés) belonging to clusters or combinations of clusters** (`releves_in_clusters()`).
- Finding **clusters that contain a given species (or set of species)** (`clusters_with_species()`).
- **Distances and relationships between clusters** based on plot-membership containment or direct plot co-membership φ (`cluster_dist()`).
- **Assignment of plots (relevés) to candidate vegetation units** using cover- and φ-based strategies (`assign_releves()`).

***

## Background

The *Cocktail* method (Bruelheide 2000, 2016) identifies sets of species that co-occur more often than expected by chance and merges them hierarchically according to the **phi coefficient of association**. Each resulting cluster is characterized by its constituent species and a threshold (*m*) indicating how many species from that cluster a plot must contain to belong to it.

For details, see the original works:

- Bruelheide, H. (2000). *A new measure of fidelity and its application to defining species groups.* **Journal of Vegetation Science**, 11, 167–178. <https://doi.org/10.2307/3236796>  
- Bruelheide, H. (2016). *Cocktail clustering – a new hierarchical agglomerative algorithm for extracting species groups in vegetation databases.* **Journal of Vegetation Science**, 27(6), 1297–1307. <https://doi.org/10.1111/jvs.12454>

***

## Installation

The development version of `cocktailr` can be installed from GitHub:

```r
install.packages("remotes")

remotes::install_github("dvynokur/cocktailr")
library(cocktailr)
```

For reproducible analyses based on a tagged release, install a specific version, for example:

```r
remotes::install_github("dvynokur/cocktailr@v0.1.0")
```
This command will work after the corresponding GitHub release/tag has been created.

***

## Typical workflow

`cocktail_cluster()` supports both **wide** (plots × species) and **long** (plot–species–value) input formats. The main example below uses a wide matrix; a compact long-format example is provided afterward.

A small end-to-end example on a toy **plots × species** matrix, showing:

1. Cocktail clustering  
2. Dendrogram plotting  
3. Selection of clusters at a φ cut **or** selection of strongest clusters by score  
4. Cluster inspection utilities  
5. Species lists  
6. Distances between clusters  
7. Plot assignment to candidate vegetation units  

```{r typical-workflow}
library(cocktailr)

# Toy plots × species matrix with percentage cover
vm <- matrix(
  c(
    60,50,40,30,  5, 0,10, 0,
    55,45,35,25, 10, 5, 0, 0,
    50,40,30,20,  5,10, 0, 5,
    45,35,25,15,  0, 5, 5, 0,
    10, 5, 0, 0, 60,50,40,30,
     5,10, 0, 0, 55,45,35,25,
     0, 5,10, 0, 50,40,30,20,
     0, 0, 5,10, 45,35,25,15
  ),
  nrow = 8, byrow = TRUE,
  dimnames = list(
    paste0("plot", 1:8),
    paste0("sp",   1:8)
  )
)

# 1) Cocktail clustering, keeping relative cover and species-cluster phi
res <- cocktail_cluster(
  vegmatrix           = vm,
  progress            = FALSE,
  plot_values         = "rel_cover",
  species_cluster_phi = TRUE,
  save_vegmatrix      = TRUE   # default; needed later by assign_releves()
)

names(res)
```

`assign_releves()` reads the vegetation matrix from `res$vegmatrix`, so keep `save_vegmatrix = TRUE` (default) if you plan to use assignment later.

For very large datasets, you can reduce memory usage by setting `save_vegmatrix = FALSE` in `cocktail_cluster()`. In that case, however, `assign_releves()` will not work unless you recompute the Cocktail object with `save_vegmatrix = TRUE`.

### Long-format input

`cocktail_cluster()` also supports **long-format** vegetation tables when `input_format = "long"`.

The default long-format column names are:

- `plot` — plot / relevé ID
- `species` — species name
- `value` — numeric cover / abundance value

Below, we convert the same toy matrix `vm` to long format and run `cocktail_cluster()`.

```{r toy-long-example}
library(dplyr)
library(tidyr)
library(tibble)

# Convert the toy wide matrix to long format
vm_long <- as.data.frame(vm) %>%
  rownames_to_column(var = "plot") %>%
  pivot_longer(
    cols = -plot,
    names_to = "species",
    values_to = "value"
  )

vm_long

# Run Cocktail clustering on long-format input
res_long <- cocktail_cluster(
  vegmatrix    = vm_long,
  input_format = "long",
  progress     = FALSE
)

# Same output structure as for wide input
names(res_long)
```

The returned Cocktail object also stores the internally used, aligned vegetation matrix in
`res_long$vegmatrix` (as a sparse `dgCMatrix`) by default (`save_vegmatrix = TRUE`).

When using long-format input, always set `input_format = "long"` explicitly.

If your long-format table uses different column names, provide a mapping via `long = list(...)`:

```{r toy-long-custom-cols, eval=FALSE}
res_long2 <- cocktail_cluster(
  vegmatrix    = my_long_table,
  input_format = "long",
  long = list(
    plot    = "PlotObservationID",
    species = "Harmonized_name_wfo",
    value   = "Relative_cover"
  ),
  progress     = FALSE
)
```

### Notes on long-format input

- Duplicate plot–species rows are aggregated using `sum()` and capped at 100.
- Empty plots (all zero after parsing and aggregation) are dropped automatically.
- The function issues warnings when duplicates are detected or empty plots are removed.

***

### 1) Visualize the dendrogram

```{r typical-plot}
cocktail_plot(
  x              = res,
  file           = NULL,       # plot to device
  phi_cut        = 0.25,
  label_clusters = TRUE,
  cex_species    = 0.9
)
```

***

### 2) Select clusters at a φ cut **or** select clusters by score

#### Option A: Parent clusters at a φ cut

```{r typical-clusters-at-cut}
phi_cut <- 0.25

parent_labels <- clusters_at_cut(
  x         = res,
  phi_cut   = phi_cut,
  as_labels = TRUE
)

parent_labels
```

#### Option B: Select clusters by score

```{r typical-select-clusters}
strong_labels <- select_clusters(
  x            = res,
  min_phi      = 0.20,
  min_k        = 1,
  min_n        = 1,
  min_score    = 0,
  score_method = "h_logk_logm",
  mode         = "strict",
  return       = "labels"
)

strong_labels
```

The default score is `h * log(k) * log(m)`, where `h` is the merge height, `k` is the number of species in the cluster and `m` is the Cocktail membership threshold. This is an operational ranking criterion. Use `score_method = "h_logk"` to rank clusters without the membership threshold, or `score_method = "h"` to rank clusters only by merge height.

(If you want the full score table:)

```{r typical-select-clusters-table}
strong_table <- select_clusters(
  x            = res,
  min_phi      = 0.20,
  min_k        = 1,
  min_n        = 1,
  min_score    = 0,
  score_method = "h_logk_logm",
  mode         = "strict",
  return       = "table"
)

strong_table
```

***

### 3) Cluster diagnostics (helper functions)

These two helpers allow quick inspection of clusters and combinations of clusters.

#### (a) List plots belonging to clusters or combinations of clusters

```{r typical-releves-in-clusters}
# Example: plots belonging to the first and third parent clusters
releves_A <- releves_in_clusters(res, clusters = parent_labels[1])
releves_B <- releves_in_clusters(res, clusters = parent_labels[3])

releves_A
releves_B
```

Combinations of clusters (list input): plots belonging to *any cluster in the set*:

```{r typical-releves-in-clusters-combinations}
cluster_set_example <- list(
  set1 = parent_labels[1:2]
)

releves_combined <- releves_in_clusters(res, clusters = cluster_set_example)
releves_combined
```

#### (b) Find clusters that contain a given species (or set of species)

```{r typical-clusters-with-species}
# Clusters that contain sp1 (default: only clusters with merge phi >= 0 are returned)
clusters_with_species(res, species = "sp1")

# Clusters that contain both sp1 and sp2,
# restricted to clusters with merge phi >= 0.2
cl <- clusters_with_species(
  res,
  species = c("sp1", "sp2"),
  match   = "all",
  min_phi = 0.2
)

cl

# Highlight clusters that contain both sp1 and sp2
cocktail_plot(
  x              = res,
  file           = NULL,
  clusters       = cl,
  label_clusters = TRUE,
  cex_species    = 0.9
)
```

***

### 4) Diagnostic species for selected clusters

Cluster-constituting species sets:

```{r typical-species-member}
diag_sp_member <- species_in_clusters(
  x      = res,
  labels = parent_labels
)

diag_sp_member
```

With φ-based filtering/ranking (uses `Species.cluster.phi`):

```{r typical-species-phi}
diag_sp_phi <- species_in_clusters(
  x                   = res,
  labels              = parent_labels,
  species_cluster_phi = TRUE,
  min_phi             = 0.20
)

diag_sp_phi
```

***

### 5) Distances and relationships between clusters

`cluster_dist()` computes distances between clusters from their binary
plot-membership vectors (`membership = Plot.cluster > 0`).

Two methods are available:

- `method = "containment"`: distance is `1 - max_containment`, where
  `max_containment` is the larger of the two directional containment values
  between a pair of clusters. This is the default and is useful for identifying
  nested or partly nested relationships among Cocktail groups.
- `method = "phi"`: distance is `1 - phi`, where `phi` is the phi coefficient
  between binary plot-membership vectors.

We must specify which clusters to compare, for example selected clusters:

```{r typical-cluster-dist}
clusters_for_dist <- select_clusters(
  x            = res,
  min_phi      = 0.20,
  min_k        = 1,
  min_n        = 1,
  min_score    = 0,
  score_method = "h_logk_logm",
  mode         = "strict",
  return       = "labels"
)

D <- cluster_dist(
  x        = res,
  clusters = clusters_for_dist,
  method   = "containment"
)

D
```

The pairwise relationship table can be inspected with:

```{r typical-cluster-dist-table}
cluster_dist(
  x        = res,
  clusters = clusters_for_dist,
  method   = "containment",
  return   = "table"
)
```

Hierarchical clustering of clusters:

```{r typical-cluster-dist-hclust}
hc_clusters <- hclust(D, method = "average")

plot(hc_clusters, main = "Cluster dendrogram (containment distance)", cex = 0.7)

grp_clusters <- cutree(hc_clusters, k = 2)
table(grp_clusters)

# Cluster combinations: group -> vector of cluster labels ("c_12", ...)
cluster_groups <- split(names(grp_clusters), grp_clusters)
cluster_groups
```

***

### 6) Visualize grouped clusters on the Cocktail dendrogram

You can pass combinations of clusters (`cluster_groups`) to `cocktail_plot()`:

```{r typical-cocktail-groups}
cocktail_plot(
  x              = res,
  clusters       = cluster_groups,
  label_clusters = TRUE,
  cex_species    = 0.9
)
```

***

### 7) Assign plots (relevés) to candidate vegetation units

`assign_releves()` assigns each plot to one of the provided candidate vegetation units using the vegetation matrix stored inside the Cocktail object (`x$vegmatrix`). This matrix is created by `cocktail_cluster()` when `save_vegmatrix = TRUE` (default). If the Cocktail object was created with `save_vegmatrix = FALSE`, `assign_releves()` will stop with an error.

Strategies:

- `"count"`     – number of candidate species present
- `"cover"`     – summed cover of candidate species
- `"phi"`       – sum of φ weights over candidate species
- `"phi_cover"` – sum of cover × φ over candidate species

Candidate species logic:

- Candidate species are the cluster-constituting species from `Cluster.species`.
- For `"phi"` and `"phi_cover"`:
  - `min_phi = NULL` (default): candidate species are the cluster-constituting species.
  - `min_phi = 0.2` (example): candidate species are taken from the **full fidelity profiles**
    (`Species.cluster.phi`), using species with φ ≥ `min_phi`.

Membership restriction:

- `plot_membership = TRUE` (default): only candidate vegetation units for which the plot is already a Cocktail member compete.
- `plot_membership = FALSE`: all candidate vegetation units compete for all plots.

Example: assign plots to the parent clusters at `phi_cut = 0.25`.  
The assignment `"+"` appears when a plot has an unresolved tie (two or more candidate vegetation units share the same best score):

```{r typical-assign}
assign_count <- assign_releves(
  x               = res,
  strategy        = "count",
  clusters        = parent_labels,
  plot_membership = TRUE,
  min_group_size  = 1L
)

table(assign_count)
```

Example: assign to candidate vegetation units represented by cluster combinations defined from containment-distance grouping:

```{r typical-assign-combined}
assign_combined <- assign_releves(
  x               = res,
  strategy        = "phi_cover",
  clusters        = cluster_groups,
  plot_membership = TRUE,
  min_phi         = 0.20,
  min_group_size  = 1L
)

table(assign_combined)
```
The returned object is a named character vector (names = plot IDs). Each value is:

- a candidate vegetation unit label such as `"u_5"` for a unit defined by a single cluster, or `"u_5_12"` for a unit represented by several clusters;
- `"+"` when there is an unresolved tie between candidate vegetation units;
- `"-"` for candidate vegetation units that were collapsed because they contain fewer plots than `min_group_size`;
- `NA` when no candidate vegetation unit wins for that plot (all scores are 0 or no eligible unit exists).

***

### (Optional) Attach assignments to a header data frame

The following example creates a simple header table with plot IDs and adds multiple assignment strategies as new columns:

```{r attach-assignments, eval=TRUE}
library(dplyr)

# Example header table with plot IDs matching the names of rel_assigned
hea <- data.frame(
  releve_number = rownames(vm)
)

strategies <- c("count", "cover", "phi", "phi_cover")

hea2 <- hea

# Choose one candidate-unit definition:
# parent_labels <- clusters_at_cut(res, phi_cut = 0.25, as_labels = TRUE)

for (s in strategies) {
  rel_assigned <- assign_releves(
    x               = res,
    strategy        = s,
    clusters        = parent_labels,
    plot_membership = TRUE,
    min_phi         = 0.20,
    min_group_size  = 2
  )

  # Align by releve_number using the names of rel_assigned
  idx <- match(as.character(hea2$releve_number), names(rel_assigned))

  colname <- paste0("unit_", s)
  hea2[[colname]] <- rel_assigned[idx]
}

dplyr::glimpse(hea2)
```

***

## Function help

See function help for details:

- `?cocktail_cluster` – build the Cocktail tree, optionally with species–cluster phi
- `?cocktail_plot` – draw dendrograms (PDF/PNG or current device)
- `?clusters_at_cut` – parent clusters at a phi cut
- `?select_clusters` – select strong clusters by score
- `?species_in_clusters` – diagnostic species for clusters or combinations of clusters
- `?releves_in_clusters` – list plots belonging to clusters or combinations of clusters
- `?clusters_with_species` – find clusters containing species
- `?cluster_dist` – distances between clusters (direct co-membership φ)
- `?assign_releves` – assign plots to candidate vegetation units using covers and φ

***

© 2026 Denys Vynokurov & Helge Bruelheide. Licensed under MIT.
